---
title: "Quality scores"
subtitle: "Code to make the plots"
author: "David Ho"
date: "9/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

#### Python script

The following is the python script to parse through multuple, gzipped FASTQ files. The output is the mean quality score at each base position for each file as well as the frequencies of the mean quality score for a read.

```
#!/usr/bin/env python

## Takes in multiple FASTQ files and returns metrics for the quality score at each bp position. Output is a tab-delim .txt file per FASTQ input

## IMPORTS

import numpy as np
import argparse
import zipfile
import gzip

## BACKGROUND DEFINITIONS

## Definte a function that gets a quality score
def convert_phred(letter):                                   
    """Converts a single ASCII character into a quality score, assuming Phred+33"""
    qscore = ord(letter) - 33   # ord takes the ASCII char and turns it into a Phred score 
    return qscore

## Arg parsing
def qual_score():
	parser = argparse.ArgumentParser(description='Fastq files past through will output a .txt file with metrics for the quality score at each bp position')
	parser.add_argument("-f", help="FASTQ file", required=True, type=str, nargs='*')
	return parser.parse_args()

args = qual_score()

## MAIN 

for fastq in args.f:
    with gzip.open(fastq,"rt") as file:      ## open a fastq file
        NR = 0
        for line in file:            
            NR += 1
            line = line.strip('\n')               # strip off \n            
            if NR == 4:                           # look at the forth line
                sequence_length = len(line)       # how many
                break
    with gzip.open(fastq,"rt") as file:       # open file
        NR = 0
       	reads = 0
        all_scores = np.zeros(sequence_length)    # running total of means
        score_freq = {}
        
        for line in file:            
            NR += 1
            line = line.strip('\n')               # strip off \n            
            if NR%4==0:                           # for every 4th line                   
                scores = np.zeros(sequence_length)         # quality scores for that read
                i = 0
                reads += 1
                
                for bp in line:
                    all_scores[i] += convert_phred(line[i]) 
                    scores[i] += convert_phred(line[i]) 
                    i += 1  
                
                read_mean = sum(scores)/len(scores)
                
                if read_mean in score_freq:
                    score_freq[read_mean] += 1
                else:
                    score_freq[read_mean] = 1

        for i in range(len(all_scores)):
            all_scores[i] = all_scores[i]/reads

        with open(fastq+"_mean_score_bp.txt",'w')	as fh: # write out the metrics to a new file
            fh.write("Mean Quality Score"+"\n")	  #Print the header you want
            i = 0
            for x in all_scores:
                fh.write(str(all_scores[i])+"\n")
                i += 1
        with open(fastq+"_freq.txt", "w") as fh:
            fh.write("Average qual score & Frequency"+"\n")
            for key,value in score_freq.items():
                fh.write(str(key)+"\t"+str(value)+"\n")

print("Created file(s) with mean quality scores at each bp position & frequency of avg quality score/read")
```


#### Import the data into R

```{r}
R1_means = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R1_means.txt", header=TRUE)
R1_freq = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R1_freq.txt", row.names = NULL, col.names = c("Mean", "Frequency"))

R2_means = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R2_means.txt", header=TRUE)
R2_freq = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R2_freq.txt", row.names = NULL, col.names = c("Mean", "Frequency"))

R3_means = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R3_means.txt", header=TRUE)
R3_freq = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R3_freq.txt", row.names = NULL, col.names = c("Mean", "Frequency"))

R4_means = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R4_means.txt", header=TRUE)
R4_freq = read.delim("/Users/davidho/Desktop/turn_in/quality_scores_output_files//R4_freq.txt", row.names = NULL, col.names = c("Mean", "Frequency"))
```

### At each bp postion for each read, what's the average quality score?
```{r}
plot(R1_means[,1],
     type="n", 
     ylim=c(30,40),
     ylab="Mean quality score",
     xlab="Bp position",
     main="Mean quality score for sequence reads")
lines(R1_means, col="blue")
lines(R4_means, col="darkgreen")
legend(87, 32, legend=c("R1", "R4"), col=c("blue", "darkgreen"), lty=1)

plot(R3_means[,1], 
     type="n", 
     ylim=c(30,40),
     ylab="Mean quality score",
     xlab="Bp position",
     main="Mean quality score for barcode reads")
lines(R2_means, col="red")
lines(R3_means)
legend(7, 32, legend=c("R2", "R3"), col=c("red", "black"), lty=1)
```

### For each file, what's the distribution of avergae quality score/read?

```{r}

par(las=1)

R1_freq$bin = cut(R1_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R1_new = aggregate(Frequency ~ bin, data=R1_freq, sum)
barplot(R1_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="lightblue",
        main="Distribution of avg read quality scores (R1)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R2_freq$bin = cut(R2_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R2_new = aggregate(Frequency ~ bin, data=R2_freq, sum)
barplot(R2_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="lightgreen",
        main="Distribution of avg read quality scores (R4)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R2_freq$bin = cut(R2_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R2_new = aggregate(Frequency ~ bin, data=R2_freq, sum)
barplot(R2_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="darkred",
        main="Distribution of avg read quality scores (R2)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R3_freq$bin = cut(R3_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R3_new = aggregate(Frequency ~ bin, data=R3_freq, sum)
barplot(R3_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="grey",
        main="Distribution of avg read quality scores (R3)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)
```

<br>

#### Look at the histograms side by side
```{r}

par(las=1, mfrow=c(2,2), cex.main=0.75)

R1_freq$bin = cut(R1_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R1_new = aggregate(Frequency ~ bin, data=R1_freq, sum)
barplot(R1_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="lightblue",
        main="Distribution of avg read quality scores (R1)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R2_freq$bin = cut(R2_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R2_new = aggregate(Frequency ~ bin, data=R2_freq, sum)
barplot(R2_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="lightgreen",
        main="Distribution of avg read quality scores (R4)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R2_freq$bin = cut(R2_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R2_new = aggregate(Frequency ~ bin, data=R2_freq, sum)
barplot(R2_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="darkred",
        main="Distribution of avg read quality scores (R2)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)

R3_freq$bin = cut(R3_freq$Mean,breaks = c(10,15,20,25,30,35,41))
R3_new = aggregate(Frequency ~ bin, data=R3_freq, sum)
barplot(R3_new$Frequency, 
        space=0, 
        width=1, 
        xlim=c(0,6),
        ylim=c(0,3.5e+08),
        ylab="Count",
        xlab="Average quality score/read",
        col="grey",
        main="Distribution of avg read quality scores (R3)",
        cex.axis = 0.5
)
axis(1, at = 0:6, labels = c(10,15,20,25,30,35,41), cex.axis=0.6)
```
